---
title: "Introduction to costnp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to costnp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(costnp)
set.seed(1)
```

## The Neyman-Pearson framework

In standard binary classification the goal is to minimize overall error. But
many real problems are asymmetric: a false negative (missing a fraud case,
failing to detect a disease) may be far more costly than a false positive.

The **Neyman-Pearson (NP) framework** addresses this directly. Instead of
minimizing total error, it asks:

> *Find a classifier that minimizes the type II error (false negative rate)
> subject to the type I error (false positive rate) being bounded above by
> $\alpha$, with high probability $1 - \delta$.*

`costnp` implements this via cost-sensitive learning: it searches over a grid
of cost parameters, fits a classifier at each value, and uses a Beta
confidence bound on the calibration set to select the cost whose classifier
provably satisfies the NP constraint.

## Setting up data

The package expects data as a list with two elements:

- `x` — a numeric feature matrix (observations × features)
- `y` — an integer vector of 0/1 labels (0 = negative class, 1 = positive class)

The NP guarantee controls the **type I error on class 0**, so label assignment
matters: class 0 is the class whose false positive rate you want to bound.

```{r data}
# Class 0: N(0, I);  Class 1: N(c(1.5, 0, 0), I)
x0 <- matrix(rnorm(200 * 3),             200, 3)
x1 <- matrix(rnorm(200 * 3, mean = 1.5), 200, 3)
data <- list(x = rbind(x0, x1), y = c(rep(0L, 200), rep(1L, 200)))
table(data$y)

# Population data — same distribution, larger sample
x0p <- matrix(rnorm(5000 * 3),             5000, 3)
x1p <- matrix(rnorm(5000 * 3, mean = 1.5), 5000, 3)
pop_data <- list(x = rbind(x0p, x1p), y = c(rep(0L, 5000), rep(1L, 5000)))
```

## Basic usage

`select_np_cost()` is the core function. The key arguments are `alpha`
(the type I error target) and `method` (the base classifier).

```{r basic}
result <- select_np_cost(
  data        = data,
  population_data = pop_data,
  alpha       = 0.1,
  delta       = 0.1,
  method      = "LR",
  seed        = 42
)
```

### Interpreting the output

```{r output}
result$cost              # selected cost parameter
result$population_type1  # type I error on population_data
result$population_type2  # type II error on population_data
result$n_calib           # calibration set size used
result$m_errors_in_calib # class-0 errors on calibration set
```

The guarantee is: with probability at least $1 - \delta = 0.9$, the
population type I error is at most $\alpha = 0.1$. The tradeoff is that
a tighter type I bound generally increases the type II error.

## `costnp` vs `costnp_plus`

The package provides two named wrappers that differ in how the final cost is
chosen once the algorithm finds the boundary between under- and
over-estimating the target:

- **`costnp()`** snaps to the last grid point that still satisfies the
  bound — the more conservative choice.
- **`costnp_plus()`** linearly interpolates between the last under-estimate and
  first over-estimate — slightly less conservative, potentially lower type II
  error.

```{r compare}
r_strict <- costnp(data, pop_data, method = "LR", seed = 42)
r_interp <- costnp_plus(data,   pop_data, method = "LR", seed = 42)

data.frame(
  method     = c("costnp", "costnp_plus"),
  cost       = c(r_strict$cost, r_interp$cost),
  type1_pop  = c(r_strict$population_type1, r_interp$population_type1),
  type2_pop  = c(r_strict$population_type2, r_interp$population_type2)
)
```

In practice the difference is small — typically one grid step — but `costnp_plus`
is preferable when minimising type II error is important and the slight
reduction in conservativeness is acceptable.

## Switching classifiers

The `method` argument selects the base classifier. All methods listed below
use stratified resampling (the default `classify_fun`) to inject the
cost-sensitivity:

| `method`   | Classifier              | Package       |
|------------|-------------------------|---------------|
| `"LR"`     | Logistic regression     | base R        |
| `"penLR"`  | Lasso logistic          | glmnet        |
| `"ridgeLR"`| Ridge logistic          | glmnet        |
| `"RF"`     | Random forest           | randomForest  |
| `"NB"`     | Naive Bayes             | naivebayes    |
| `"LDA"`    | Linear discriminant     | MASS          |
| `"SVM"`    | Support vector machine  | e1071         |

```{r methods, eval = FALSE}
r_rf <- select_np_cost(data, pop_data, method = "RF", seed = 42)
r_lda <- select_np_cost(data, pop_data, method = "LDA", seed = 42)
```

## Threshold sweep instead of resampling

An alternative to resampling is to fit a single unweighted classifier and
sweep the decision threshold. Use `classify_fun_threshold` paired with
`threshold_fun = function(cost) cost`:

```{r threshold}
r_thresh <- select_np_cost(
  data            = data,
  population_data = pop_data,
  method          = "LR",
  classify_fun   = classify_fun_threshold,
  threshold_fun  = function(cost) cost,
  cost_start     = 0.99,
  cost_end       = 0.50,
  seed           = 42
)
r_thresh$cost            # now interpreted as a decision threshold
r_thresh$population_type1
```

## Custom classifiers

If none of the built-in methods fit your needs, you can supply your own
`classify_fun`. It must have the signature:

```r
function(xtrain, ytrain, cost, xnew, method = NULL, ...)
```

and return `list(prob, obj)` where `prob` is a numeric vector of predicted
probabilities for `xnew` and `obj` is the fitted model. When using a custom
`classify_fun`, `method` can be left as `NULL`.

```{r custom, eval = FALSE}
my_classifier <- function(xtrain, ytrain, cost, xnew, method = NULL, ...) {
  # example: weight observations by cost
  w <- ifelse(ytrain == 0, cost, 1 - cost)
  df <- data.frame(y = ytrain, xtrain)
  fit <- glm(y ~ ., data = df, family = binomial(), weights = w)
  prob <- predict(fit, newdata = data.frame(xnew), type = "response")
  list(prob = prob, obj = fit)
}

r_custom <- select_np_cost(
  data            = data,
  population_data = pop_data,
  classify_fun    = my_classifier,
  seed          = 42
)
```
